---
title: "Bulk RNAseq_DESeq2"
author: Ludmila Danilova (modified by Daniel Shu)
date: 2021-11-09 (2022-05-11) with further modifications on 2022/10/5
output:
  html_document:
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE
)
```

# I. Setup
### a. Load libraries

```{r setup, include=FALSE}
library(tidyverse)
library(ggpubr)
library(DESeq2)
library(ggrepel)
library(msigdbr)
library(data.table)
library("DT")
library(RColorBrewer)
library(viridis)
library(EnhancedVolcano)
library(ggplotify)
library(tools)
library(pca3d)
library(edgeR)
library(pheatmap)
library("ComplexHeatmap")
library(circlize)
library(ggprism)
library(writexl)

dataDir = "data/BulkRNAseq/rna_gene_expression/"
output.path = "output/RNAseq/"
ifelse(!dir.exists(output.path), dir.create(output.path), paste0(output.path, " ", "directory already exists"))
```

# II. Load data
### a. load sampInfo file
```{r}
sampInfo <- readRDS("output/sampInfo.rds") 

# fix patient IDs to map to Personalis folder names
sampInfo$Publication.ID = sampInfo$Publication.ID %>% 
  str_replace_all(c(#"17136"="J17136","-"="_",
                                        "OT1" = "OT1_Tumor",
                                        "OT2" = "OT2_Tumor_Replacement",
                                        "OT4"= "OT4_Tumor",
                                        "OT5"= "OT5_Tumor_Replacement",
                                        "OT6"="OT6_Tumor"))

rownames(sampInfo) = sampInfo$Publication.ID

# sampInfo %>% View
sampInfo <- sampInfo %>%  
  filter(group=="Neoadjuvant")  #subset for only neoadjuvant treated tumors

sampInfo <- sampInfo %>% filter(!(BulkRNA.seq=="0")) %>% #subset for samples with bulk RNAseq data
  arrange(Publication.ID) #reorder to facilitate matching up with patient.names object in next code chunk
  
#add TLS column
TLS.mean = mean(sampInfo$TLS.Density)
sampInfo$TLS = if_else(sampInfo$TLS.Density < TLS.mean, "Low", "High")
sampInfo$TLS = as.factor(sampInfo$TLS)
sampInfo$TLS = factor(sampInfo$TLS, levels=c("Low", "High"))

#Set colors for analysis
TLSCol = c('High' =  "#631879E5", 'Low' =  "#008280E5")
Etiology.2Col = c('Viral' = '#36454F', 'Non-Viral' = '#D3D3D3')
respCol=c('MPR/pCR'="#EE0000E5", 'pPR' = "#008B45E5", 'NR' = "#3B4992E5")
treatmentCol =  c('anti-PD1 + TKI' = "#0056a2", 
                  # 'anti-PD1 + anti-CTLA4 + TKI' = "#56a200",
                  'anti-PD1'= "#A20056E5"
                    )
relapseCol =   c(Yes="#d6d091", 
                 No="#91bad6") 
TLSdensityCol = colorRamp2(c(0,0.4), c("#e9e6f3", "#631879E5"))
```


### c. Read in count data for all patients and create tpmData and countData objects

```{r echo=FALSE,  fig.width = 10}
tpmData = c()
countData = c()
files = list.files(dataDir, pattern = "_rna_gene_expression_report.tsv", full.names = T, recursive=T)
for( i in files){
   dat = read.table(i,
                   header = T, 
                   stringsAsFactors = F, 
                   as.is = T, 
                   fill = T, 
                   comment.char = "", 
                   sep = '\t')
  tpmData = cbind(tpmData,dat[which(!duplicated(dat$Gene.Symbol)),'TPM'])
  countData = cbind(countData,dat[which(!duplicated(dat$Gene.Symbol)),'RNA.Seq.Raw.Counts'])
}

# extract names from files vector, changing strings to match Publication.ID column of object sampInfo
patient.names <- file_path_sans_ext(basename(files))%>%
  str_replace(".*RNA_", "") %>%
  str_replace("_FFPE.*", "") %>%
  str_replace("_tumor.*", "") %>%
  str_replace("-", "_") %>% sort
patient.names
#doublecheck that patient names object contains the correct information. this should return TRUE TRUE TRUE, etc.
patient.names == sampInfo$Publication.ID

#assign rownames to tpmData and countData as gene symbols 
rownames(tpmData) = rownames(countData) = dat[which(!duplicated(dat$Gene.Symbol)),'Gene.Symbol']
#assign colnames to tpmData and countData using modified filenames
colnames(tpmData) =colnames(countData) = patient.names

```

### d. Check counts data using log2 transformation. Here we add “+1” because log0 does not exist but log1=0.

```{r echo=FALSE,  fig.width = 10}
counts.matrix.log <- log2(countData+1)
par(mar=c(12,5,2,2))
boxplot(counts.matrix.log, las=2, ylab="log2(counts+1)")
rm(counts.matrix.log)

# based on this, I ultimately excluded OT2_Tumor_Replacement and OT6_Tumor from the analysis since the median count was 0. 
# first removed them from sampInfo 
sampInfo <- sampInfo %>%  filter(!(Publication.ID==c("OT2_Tumor_Replacement", "OT6_Tumor")))
#then removed them from countData and tpmData objects
countData = countData[, sampInfo[,1]]
tpmData = tpmData[, sampInfo[,1]]

#make new plot of log2 transformation again with OT2 and OT6 excluded
counts.matrix.log <- log2(countData+1)
par(mar=c(12,5,2,2))
boxplot(counts.matrix.log, las=2, ylab="log2(counts+1)")
rm(counts.matrix.log)
```

# III. DESeq2 analysis of TLS high vs TLS low
### a. Preprocessing: Create DeSeq object from counts matrix and do PCA analysis
### note here I did not include batch as variate based on results of plotPCA below

```{r echo=FALSE,  fig.width = 10}
dds <- DESeqDataSetFromMatrix(countData, 
                              sampInfo, 
                              design = ~TLS)
dds <- dds[rowSums(counts(dds))>1,]

# vst transformation for visualization
logSTCDataSet <- vst(dds)

# expression matrix
expr = assays(logSTCDataSet)[[1]]

par(mar=c(12,5,2,2))
boxplot(expr, las=2, ylab="expression matrix after vst")

### PCA using plotPCA function in DESeq2. 
pcaData <- plotPCA(logSTCDataSet, intgroup = "TLS",returnData=T)
percentVar <- round(100*attr(pcaData,"percentVar"))
pcaData$TLS <- factor(pcaData$TLS, levels=c("High", "Low"))
ggplot(pcaData,aes(PC1,PC2, fill=TLS))+#, label=name))+
  geom_point(shape=21,size=9,colour="black")+
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()+
  scale_fill_manual("",values = TLSCol,labels=c("High"="TLS High", "Low"="TLS Low"))+
  ggprism::theme_prism()+
  theme(legend.position = c(0.5,0.96), legend.text=element_text(face="bold"),
        text = element_text(size=16)
        )+
  guides(fill = guide_legend(nrow = 1))

ggsave("output/RNAseq/pca_by_TLS.pdf",  bg = "transparent", height=6, width=9)

# This identifies J17136_P13 and OT4_Tumor as outliers
plotPCA(logSTCDataSet, intgroup = "Batch") + geom_text(aes(label=name),hjust=0, vjust=0) +
  coord_cartesian(clip="off")
ggsave("output/RNAseq/pca_by_Batch.pdf")

# Second check using PCA3d library
# This identifies OT4_Tumor and J17136_P13 as outliers. OT5_Tumor_Replacement and J17136_02 are also outliers, but they are at least on the same PC 2 scale as the other samples.  
# Overall, no clear batch to batch differences, so did not exclude samples based on batch effect
pcs <- prcomp(t(expr),scale=T)

pca2d(prcomp(t(expr), scale=T), 
      group=sampInfo$Batch,
      legend="top",
      show.labels=T
      )
title('PCA of vst counts orig filter')
ggsave("output/RNAseq/2D_pca_by_Batch.pdf")

```

###  b. Perform DEseq2 analysis
```{r echo = FALSE}
dds_res <- DESeq(dds, betaPrior = F)
resultsNames(dds_res)
DE_res = results(dds_res)

# Determine DEGs for visualization below
FCthr = 1
FDR = 0.05
plot_FCthr = 10
plot_FDR = 1e-30

DE_res = DE_res[which(!is.na(DE_res$padj)),]
res <- list(DE_res = DE_res, 
		# gene names to plot on volcano
		signPlot = row.names(DE_res)[which(abs(DE_res$log2FoldChange) > plot_FCthr | DE_res$padj < plot_FDR)],
		# significant genes to save into file
		signSave = row.names(DE_res)[which(abs(DE_res$log2FoldChange) > FCthr & DE_res$padj < FDR)])
	
#	print(c('genes to print',length(res$signPlot)))
	print(c('significant genes',length(res$signSave)))
	# gene names to plot on volcano
	DE_res$label = F
	DE_res[res$signPlot,'label'] = TRUE
	DE_res$gene_symbol = rownames(DE_res)
	DE_res$diffexpressed <- FALSE
	DE_res[res$signSave,'diffexpressed'] <- TRUE

significantGenes <-	as.data.frame(res$DE_res[res$signSave,])	
write.csv(significantGenes, file = "./output/RNAseq/significantGenes.csv", quote = F)
datatable(significantGenes)
```
### c. Visualization
#### 1. Volcano plots
```{r volc_plots, eval=T}
#This was Luda's code
pdf("output/RNAseq/volcano_plots_luda.pdf", width=10, height=8)
	gobj = ggplot(data=data.frame(DE_res), aes(x=log2FoldChange, y=-log10(padj), col=diffexpressed)) + 
		geom_point() + 
		theme_minimal() +		
		geom_text_repel(aes(x = log2FoldChange, y = -log10(padj), label = ifelse(label == T, rownames(DE_res),""))) +
	   xlab("log2 fold change") + 
	  ylab("-log10 adjusted p-value")

print(gobj)
dev.off()

#Make a volcano plot using enhancedVolcano
volc_plot <- EnhancedVolcano(DE_res,
    lab = rownames(DE_res),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'TLS high vs low', 
    subtitle = NULL,
    caption = paste0("total = ", nrow(DE_res), " genes"),
  xlim = c(min(DE_res[['log2FoldChange']], na.rm = TRUE) - 0.5, max(DE_res[['log2FoldChange']], na.rm = TRUE) +
    0.5),
  ylim = c(0, max(-log10(DE_res[['pvalue']]), na.rm = TRUE) + 0.5),
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 1,
    labSize = 3.25,
    labCol = 'black',
    labFace = 'bold',
    colAlpha=1,
    legendLabSize = 12,
    legendIconSize = 3,
    gridlines.major = FALSE, gridlines.minor = FALSE
)

pdf("output/RNAseq/volcano_plots_1.pdf", width=10, height=8)
volc_plot
dev.off()

#volcano plot for certain goi
#load volcano_goi object, which contains numerous immune related genes
source("./scripts/volcano_goi.R")

volc_plot.2 <- EnhancedVolcano(DE_res,
    lab = rownames(DE_res),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'TLS high vs low', 
    subtitle = 'Curated TLS- or B- cell relevant list',
    caption = paste0("total = ", nrow(DE_res), " genes"),
    # xlim = c(-10, 10),
    # ylim = c(0, 20),
      xlim = c(min(DE_res[['log2FoldChange']], na.rm = TRUE) - 0.5, max(DE_res[['log2FoldChange']], na.rm = TRUE) +
    0.5),
  ylim = c(0, max(-log10(DE_res[['pvalue']]), na.rm = TRUE) + 0.5),
    selectLab = volcano_goi[volcano_goi %in% rownames(significantGenes)], 
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 0.5,
    labSize = 3,
    labCol = 'black',
    labFace = 'bold',
    boxedLabels = TRUE,
    colAlpha=1,
    legendLabSize = 12,
    legendIconSize = 3,
    maxoverlapsConnectors = 50,
    drawConnectors = TRUE,
    widthConnectors = 0.8,
    typeConnectors = "closed",
    arrowheads = FALSE,
    gridlines.major = FALSE, gridlines.minor = FALSE
)
pdf("output/RNAseq/volcano_plots_2.pdf", width=10, height=8)
volc_plot.2
dev.off()

chemo12 = unlist(read.table(paste0(dataDir,'chemokine_signature.txt')))

volc_plot.chemo12 <- EnhancedVolcano(DE_res,
    lab = rownames(DE_res),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'TLS high vs low', 
    subtitle = '12-chemokine gene signature',
    caption = paste0("total = ", nrow(DE_res), " genes"),
      xlim = c(min(DE_res[['log2FoldChange']], na.rm = TRUE) - 0.5, max(DE_res[['log2FoldChange']], na.rm = TRUE) +
    0.5),
  ylim = c(0, max(-log10(DE_res[['pvalue']]), na.rm = TRUE) + 0.5),
    selectLab = chemo12,
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 1,
    labSize = 3.25,
    labCol = 'black',
    labFace = 'bold',
    colAlpha=1,
    legendLabSize = 12,
    legendIconSize = 3,
    maxoverlapsConnectors = 50,
    drawConnectors = TRUE,
    widthConnectors = 0.8,
    typeConnectors = "closed",
    arrowheads = FALSE,
    gridlines.major = FALSE, gridlines.minor = FALSE
)

pdf("output/RNAseq/volcano_plots_chemo12.pdf", width=10, height=8)
volc_plot.chemo12
dev.off()

volc_plot_select <- EnhancedVolcano(DE_res,
    lab = rownames(DE_res),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'TLS high vs low', 
    subtitle = NULL,
    caption = paste0("total = ", nrow(DE_res), " genes"),
     xlim = c(min(DE_res[['log2FoldChange']], na.rm = TRUE) - 0.5, max(DE_res[['log2FoldChange']], na.rm = TRUE) +
    0.5),
  ylim = c(0, max(-log10(DE_res[['pvalue']]), na.rm = TRUE) + 0.5),

   selectLab = c("MS4A1", "TNFRSF13C", "TNFRSF13B", "IL6", "CCL2", "IL7R", "CTLA4", "IL10", "CCR7", "CD28", "GPR183", "CD79B", "CXCR5", "FCRL1", "CD79A", "CCL19", "TNFSF13B", "CD69", "IL17C", "ENPTD1", "MADCAM1"), #comment this line out to get all genes labeled
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 1,
    labSize = 3,
    labCol = 'black',
    labFace = 'bold',
    boxedLabels = TRUE,
    colAlpha=1,
    legendLabSize = 12,
    legendIconSize = 3,
    maxoverlapsConnectors = 100,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    typeConnectors = "closed",
    arrowheads = FALSE,
    gridlines.major = FALSE, gridlines.minor = FALSE
)

pdf("output/RNAseq/volcano_plots_selectGenes.pdf", width=10, height=8)
volc_plot_select
dev.off()

volc_plot_select.widerYAxis <- EnhancedVolcano(DE_res,
    lab = rownames(DE_res),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'TLS high vs low', 
    subtitle = NULL,
    caption = paste0("total = ", nrow(DE_res), " genes"),
     xlim = c(min(DE_res[['log2FoldChange']], na.rm = TRUE) - 0.5, max(DE_res[['log2FoldChange']], na.rm = TRUE) +
    0.5),
  ylim = c(0, max(-log10(DE_res[['pvalue']]), na.rm = TRUE) + 0.5),

   selectLab = c("MS4A1", "TNFRSF13C", "TNFRSF13B", "IL6", "CCL2", "IL7R", "CTLA4", "IL10", "CCR7", "CD28", "GPR183", "CD79B", "CXCR5", "FCRL1", "CD79A", "CCL19", "TNFSF13B", "CD69", "IL17C", "ENPTD1", "MADCAM1"), #comment this line out to get all genes labeled
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 1,
    labSize = 3,
    labCol = 'black',
    labFace = 'bold',
    boxedLabels = TRUE,
    colAlpha=1,
    legendLabSize = 12,
    legendIconSize = 3,
    maxoverlapsConnectors = 100,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    typeConnectors = "closed",
    arrowheads = FALSE,
    gridlines.major = FALSE, gridlines.minor = FALSE
)

pdf("output/RNAseq/volcano_plots_widerY.pdf", width=10, height=8)
volc_plot_select.widerYAxis
dev.off()


volc_plot_select.narrowXY<- EnhancedVolcano(DE_res,
    lab = rownames(DE_res),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'TLS low vs TLS high', 
    subtitle = NULL,
    caption = paste0("total = ", nrow(DE_res), " genes"),
   xlim = c(-10, 10),
   ylim = c(0, 14),
   selectLab = c("MS4A1", "TNFRSF13C", "TNFRSF13B", "IL6", "CCL2", "IL7R", "CTLA4", "IL10", "CCR7", "CD28", "GPR183", "CD79B", "CXCR5", "FCRL1", "CD79A", "CCL19", "TNFSF13B", "CD69", "IL17C", "ENPTD1", "MADCAM1"), #comment this line out to get all genes labeled
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 1,
    labSize = 5.25,
    labCol = 'black',
    labFace = 'bold',
    boxedLabels = T,
    colAlpha=1,
    legendLabSize = 12,
    legendIconSize = 4,
    maxoverlapsConnectors = 100,
    drawConnectors = TRUE,
    widthConnectors = 1,
    typeConnectors = "closed",
    arrowheads = FALSE,
    gridlines.major = FALSE, gridlines.minor = FALSE,
  cutoffLineWidth = 0.8
  )

volc_plot_select.narrowXY$labels$y <- "-log10(Padj)"
volc_plot_select.narrowXY$labels$x <-"log2 fold change"

volc_plot_select.narrowXY <- volc_plot_select.narrowXY+scale_y_continuous(expand = c(0, 0))+
  theme(plot.title = element_text(hjust=0.5),
        axis.line = element_line(linewidth=1),
        legend.position="top",
      legend.box.spacing = unit(10, "pt")) # The spacing between the plotting area and the legend box (unit)
        
volc_plot_select.narrowXY

pdf("output/RNAseq/volcano_plots_narrowXY.pdf",width=9,height=8)
volc_plot_select.narrowXY
dev.off()
```
#### 2. Heatmap for all DEGs
Heatmap will be annotated for genes present in select GOBP pathways 
```{r heatmaps, echo = FALSE, eval=T}
#scale expression data
expr_scale = t(scale(t(expr), center=T))
head(expr)
head(expr_scale)

# 1a.heatmap of significant genes with complex heatmap
mat = expr_scale[res$signSave, ]

## load T cell, B cell, and antigen presenting cell gene lists, etc 
m_df = msigdbr(species = "Homo sapiens", category = 'C5', subcategory = "GO:BP") # gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name)

df=data.frame(names=names(m_list),
              n_hits = sapply (1:length(m_list), function(i) {sum(rownames(mat) %in% m_list[[i]])})
) %>% arrange(., desc(n_hits))

write_csv(df,file="output/RNAseq/n_significantGenes_in_GOBP_pathways.csv")

cc = m_list$GOBP_T_CELL_ACTIVATION
ccl = rownames(mat) %in% cc

rp = m_list$GOBP_B_CELL_ACTIVATION
rpl = rownames(mat) %in% rp

ck = m_list$GOBP_CYTOKINE_PRODUCTION_INVOLVED_IN_IMMUNE_RESPONSE
ckl = rownames(mat) %in% ck

ap = m_list$GOBP_DENDRITIC_CELL_ANTIGEN_PROCESSING_AND_PRESENTATION
apl = rownames(mat) %in% ap

#export list of genes identified by cross-reference against GOBP
deg_by_geneSet <- list(t_cell_activation = rownames(mat)[ccl], 
       b_cell_activation = rownames(mat)[rpl],
       cytokine_production = rownames(mat)[ckl],
       antigen_presentation = rownames(mat)[apl]
       ) %>% 
  lapply(., as.data.frame) %>% lapply(., setNames, "Gene")

write_xlsx(deg_by_geneSet,path="output/RNAseq/deg_by_geneSet.xlsx")

row_ann_l <- rownames(mat) %in% c(cc, rp,  
                                  ck, 
                                  ap)



#reverse levels on sampInfo$TLS to for sake of heatmap
sampInfo$TLS <- factor(sampInfo$TLS, levels=c("High", "Low"))

#add Etiolgoy.2 column to sampInfo
sampInfo$Etiology.2 = case_when(
      str_detect(sampInfo$Etiology, regex("HCV|HBV/HCV", ignore_case=TRUE)) ~ "Viral",
      str_detect(sampInfo$Etiology, regex("Unknown|Alcohol|NASH", ignore_case=TRUE)) ~ "Non-Viral")
sampInfo$Etiology.2 = factor(sampInfo$Etiology.2,levels=c("Viral", "Non-Viral"))
      
#define heatmap annotations for top of heatmap
ha = HeatmapAnnotation(TLS= sampInfo[,'TLS'],
                       Etiology = sampInfo[,'Etiology.2'],
                       Treatment = sampInfo[, 'Treatment'],
                       Response = sampInfo[,'Response'], 
                       Relapse = sampInfo[, 'Relapse'],
                       `TLS Density` = sampInfo[,"TLS.Density"],
                       col = list(TLS=TLSCol,
                                  Etiology = Etiology.2Col,
                                  Response = respCol,
                                  Treatment=treatmentCol,
                                  Relapse = relapseCol
                                  ,
                                  `TLS Density` = TLSdensityCol
                                  ),
                       simple_anno_size = unit(0.3, "cm"),
                       annotation_name_gp= gpar(fontsize = 9, fontface="bold"),
                       annotation_name_side = "left",
                       gp = gpar(col = "white"))

library(GetoptLong)
ht_list = Heatmap(mat,
        col= inferno(100), #col_fun, 
          #column_title = qq("DEGs in TLS high vs TLS low (n = 814)"),
        # column_title_gp = gpar(fontsize = 10, fontface = "bold"),
        show_column_names=F,#,width=unit(14, "cm"),
        name = "scaled_expr",
        row_names_gp = gpar(fontsize = 4),
        top_annotation = ha, #see above
        heatmap_legend_param = list(title = "Expression"
                                    # ,direction="horizontal"
                                    ))+ 
  Heatmap(ccl + 0, name = "T cell activation", col = c("0" = "white", "1" = "red"),
        show_heatmap_legend = FALSE, width = unit(5, "mm"), 
        column_names_side = "top", column_names_rot = 90,
        column_names_gp = gpar(fontsize = 9, fontface="bold"))+
  Heatmap(rpl + 0, name = "B cell activation", col = c("0" = "white", "1" = "purple"),
        show_heatmap_legend = FALSE, width = unit(5, "mm"),
        column_names_side = "top", column_names_rot = 90,
        column_names_gp = gpar(fontsize = 9, fontface="bold"))+
  Heatmap(ckl + 0, name = "Cytokine production", col = c("0" = "white", "1" = "blue"),
        show_heatmap_legend = FALSE, width = unit(5, "mm"),
        column_names_side = "top", column_names_rot = 90,
        column_names_gp = gpar(fontsize = 9, fontface="bold"))+
  Heatmap(apl + 0, name = "Antigen presentation", col = c("0" = "white", "1" = "orange"),
        show_heatmap_legend = FALSE, width = unit(5, "mm"),
        column_names_side = "top", column_names_rot = 90,
        column_names_gp = gpar(fontsize = 9, fontface="bold"))+
  # Heatmap(lml + 0, name = "Leukocyte migration", col = c("0" = "white", "1" = "green"),
  #       show_heatmap_legend = FALSE, width = unit(5, "mm"),
  #       column_names_side = "top", column_names_rot = 90,
  #       column_names_gp = gpar(fontsize = 9, fontface="bold"))+
  rowAnnotation(link = anno_mark(at = which(row_ann_l
                                              #& base_mean > quantile(base_mean, 0.25)
                                              ),
                                   labels = rownames(mat)[row_ann_l 
                                                          # &base_mean > quantile(base_mean, 0.25)
                                                          ],
        labels_gp = gpar(fontsize = 7, fontface="italic"), padding = unit(0.2, "mm")))

ht_list = draw(ht_list, main_heatmap = "scaled_expr", ht_gap = unit(0.75, "mm"), 
               merge_legend = TRUE,
               annotation_legend_side="right",
               heatmap_legend_side="right")

pdf("output/RNAseq/complexHeatmap.pdf", width = 7, height = 16)
ht_list
dev.off()
```

#### 3. Heatmap for 12-chemokine signature

```{r}
# use the object containing scaled VST transformed expression data, expr_scale, to plot 12 chemokine signature
chemo12 = unlist(read.table(paste0(dataDir,'chemokine_signature.txt')))

mat2 = expr_scale[rownames(expr_scale) %in% chemo12,]

ht_list2 =
  Heatmap(mat2,
        col= rev(pals::brewer.rdbu(100)), #col_fun, 
          # column_title = qq("DEGs in TLS high vs TLS low (n = 814)"),
        # column_title_gp = gpar(fontsize = 10, fontface = "bold"),
        show_column_names=F,#,width=unit(14, "cm"),
        name = "scaled_expr",
        row_names_gp = gpar(fontsize = 10),
        top_annotation = ha, #see above
        heatmap_legend_param = list(title = "Expression"
                                    # ,direction="horizontal"
                                    ),
        width = ncol(mat2)*unit(8, "mm"),
    height = nrow(mat2)*unit(6, "mm")
    )

ht_list2 = draw(ht_list2, main_heatmap = "scaled_expr", ht_gap = unit(0.75, "mm"), 
               merge_legend = F,
               annotation_legend_side="bottom",
               heatmap_legend_side="right")
pdf("output/RNAseq/complexHeatmap_12chemokine.pdf")
ht_list2
dev.off()

```

# IV. Save object
```{r}
#save DE_res object for GSEA in next rmd
saveRDS(DE_res, paste0(output.path,"DE_res.rds"))
```

# V. SessionInfo

```{r echo=FALSE}
sessionInfo()
writeLines(capture.output(sessionInfo()), "sessionInfo_BulkRNAseq_DESeq2.txt")
```
