---
title: "Bulk RNAseq_GSEA"
author: Ludmila Danilova (later modified by Daniel Shu)
date: 2021-11-09 (2022-05-11) with further modifications on 2022/10/5
output:
  html_document:
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE
)
```

# I. Setup
### a. Load libraries

```{r setup, include=FALSE}
library(tidyverse)
library(ggpubr)
library(DESeq2)
library(ggrepel)
library(msigdbr)
library(data.table)
library("DT")
library(RColorBrewer)
library(viridis)
library(EnhancedVolcano)
library(ggplotify)
library(tools)
library(pca3d)
library(edgeR)
library(pheatmap)
library("ComplexHeatmap")
library(circlize)
library(ggprism)
library(writexl)
library(fgsea)

```

# II. Load data

```{r}
DE_res <- readRDS("output/RNASeq/DE_res.rds")
output.path = "./output/RNAseq_GSEA/"
ifelse(!dir.exists(file.path(output.path)), dir.create(file.path(output.path)), FALSE)
```

### e. GSEA

Create the modular GSEA analysis code (N.B. this will be rerun again below for other gene sets)

```{r gsea_run, echo=FALSE,  fig.width = 10, eval=FALSE}
#create output directory if none exists
ifelse(!dir.exists(file.path(output.path, geneSet)), dir.create(file.path(output.path, geneSet)), FALSE)

# function that uses DE results and runs gene set enrichment analysis on paths
runGSEA_padj = function(deRes,paths)
{
  # library(fgsea)
  # create a stat object to use in the pathway analysis
  stat <- deRes$stat
  names(stat) <- row.names(deRes)
  
  # run GSE
  gseTest_h = fgsea(paths,stat)
  gseTest_h = gseTest_h[order(gseTest_h$NES),]
  return(gseTest_h)
}
gseTest_h = runGSEA_padj(DE_res, m_list)
fwrite(gseTest_h, file= paste0(output.path, geneSet, '/pathwayAnalysisResults_', geneSet, '.csv'), 
       sep=",", 
       sep2=c("", " ", ""))

# datatable(gseTest_h)
```


```{r gsea_plots, echo=FALSE,  eval=FALSE}
##Plot the normalized enrichment scores. Color the bar indicating whether or not the pathway was significant:
## NOTE: This is additional code I have added to Luda's following this vignette online for fgsea (https://stephenturner.github.io/deseq-to-fgsea/#using_the_fgsea_package)


### generate plots for enrichment score
enrichment.score <- function(deRes, paths, pathway) 
{
  # library(fgsea)
  stat <- deRes$stat
  names(stat) <- row.names(deRes)

  x <- plotEnrichment(paths[[pathway]],
                 stat) + labs(title = pathway) + theme(plot.title =
                                                         element_text(hjust=0.5,face="bold",size=16),
                                                       axis.title = element_text(face="bold",size=14),
                                                       axis.text = element_text(face="bold",size=12))
  # x$layers[[1]]$aes_params$size  = 1 #this comes from https://stackoverflow.com/questions/64035941/changing-the-point-size-of-a-ggplot2-object-after-creation
  x$layers[[5]]$aes_params$size = 1
  print(x)
}

########
########
#make object up.pathways for every gene set identified with p < 0.05 and positive enrichment score
gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES>0,]$pathway -> up.pathways
#plot these
pdf(paste0(output.path, geneSet, "/gsea_enrichment_score_",geneSet,".pdf"),width=7,height=5)
for (i in up.pathways) {enrichment.score(DE_res, m_list, i)}
dev.off()
```

#### Now do the GSEAs. First gene set will be HALLMARK  (n=50)

```{r HALLMARK, echo=FALSE}
geneSet = "HALLMARK" #set name of gene set under analysis
m_df = msigdbr(species = "Homo sapiens", category = 'H') # retrieve the specific gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name)
names(m_list) = str_replace_all(names(m_list),"_"," ") %>% 
  str_replace(., paste0(geneSet, " "), "") #%>%

#gsea_run block is commented out since I've run it before, instead load the pathwayAnalysisResults file
<<gsea_run>>
gseTest_h <- read_csv(file = paste0(output.path, geneSet, "/pathwayAnalysisResults_", geneSet, ".csv"), show_col_types=F)

##############
#make gsea plots
#clean up names on plots
gseTest_h$pathway = str_replace_all(gseTest_h$pathway, "_", " ") %>%
  str_replace(., paste0(geneSet, " "), "") #%>%
gseTest_h$pathway %in% names(m_list)
#first make enrichment score plots (these need to be included in these code chunks because the size of the text changes)

pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '_1.pdf'),height=8,width=7)
ggplot(gseTest_h, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip()+
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 10,face="bold"),legend.position = "right",
        text = element_text(face ="bold"))+
  ggsci::scale_fill_npg()
dev.off()

#make plots for only significant pathways with padj<0.05
pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '_2.pdf'), height=10, width=10)
ggplot(gseTest_h[gseTest_h$padj<0.05,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 10,face="bold"))+
  scale_fill_manual(values="#355C7D")
dev.off()

#only plots w/ padj <0.05 and NES > 0
pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '_3.pdf'), height=10, width=10)
ggplot(gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,],aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 10,face="bold"))+
  scale_fill_manual(values="#355C7D")
dev.off()
#then run stock code
<<gsea_plots>>
```

#### GSEA for KEGG pathways (n=186)

```{r KEGG, echo=FALSE}
geneSet = "KEGG" #set name of gene set under analysis
m_df = msigdbr(species = "Homo sapiens", category = 'C2', subcategory = "CP:KEGG") # gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name) 
names(m_list) = str_replace_all(names(m_list),"_"," ") %>%    str_replace(., paste0(geneSet, " "), "") #%>%

#gsea_run block is commented out since I've run it before, instead load the pathwayAnalysisResults file
<<gsea_run>>
# gseTest_h <- read_csv(file = paste0(output.path, geneSet, "/pathwayAnalysisResults_", geneSet, ".csv"), show_col_types=F)

##############
#make gsea plots
#clean up names on plots
gseTest_h$pathway = str_replace_all(gseTest_h$pathway, "_", " ") %>%
  str_replace(., paste0(geneSet, " "), "") #%>%

#first make enrichment score plots (these need to be included in these code chunks because the size of the text changes)

pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '.pdf'), height=12, width=8)

ggplot(gseTest_h, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip()+
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 6))

#make plots for only significant pathways with padj<0.05
ggplot(gseTest_h[gseTest_h$padj<0.05,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 9),
        legend.position = "bottom")+
  scale_fill_manual(values="#355C7D")

#only plots w/ padj <0.05 and NES > 0
ggplot(gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,],aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 10))+
  scale_fill_manual(values="#355C7D")

dev.off()
#then run stock code
<<gsea_plots>>
```

#### GSEA for BIOCARTA gene sets (n=292)

```{r BIOCARTA, echo=FALSE}
geneSet = "BIOCARTA" #set name of gene set under analysis
m_df = msigdbr(species = "Homo sapiens", category = 'C2', subcategory = "CP:BIOCARTA") # gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name) 
names(m_list) = str_replace_all(names(m_list),"_"," ") %>%    str_replace(., paste0(geneSet, " "), "") #%>%

#gsea_run block is commented out since I've run it before, instead load the pathwayAnalysisResults file
<<gsea_run>>
# gseTest_h <- read_csv(file = paste0(output.path, geneSet, "/pathwayAnalysisResults_", geneSet, ".csv"), show_col_types=F)

##############
#make gsea plots
#clean up names on plots
gseTest_h$pathway = str_replace_all(gseTest_h$pathway, "_", " ") %>%
  str_replace(., paste0(geneSet, " "), "") #%>%
# str_to_title()

#first make enrichment score plots (these need to be included in these code chunks because the size of the text changes)

pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '.pdf'), height=10, width=10)

ggplot(gseTest_h, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip()+
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 3))

#make some plots, but only include pathways with padj<0.05
ggplot(gseTest_h[gseTest_h$padj<0.05,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 10))+
  scale_fill_manual(values="#355C7D")

#only plots w/ padj <0.05 and NES > 0
ggplot(gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,],aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 10))+
  scale_fill_manual(values="#355C7D")
dev.off()
#then run stock code
<<gsea_plots>>
```

#### GSEA for PID gene sets (n=196)

```{r PID, echo=FALSE}
geneSet = "PID" #set name of gene set under analysis
m_df = msigdbr(species = "Homo sapiens", category = 'C2', subcategory = "CP:PID") # gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name) 
names(m_list) = str_replace_all(names(m_list),"_"," ") %>%    str_replace(., paste0(geneSet, " "), "") #%>%

#gsea_run block is commented out since I've run it before, instead load the pathwayAnalysisResults file
<<gsea_run>>
# gseTest_h <- read_csv(file = paste0(output.path, geneSet, "/pathwayAnalysisResults_", geneSet, ".csv"), show_col_types=F)

##############
#make gsea plots
#clean up names on plots
gseTest_h$pathway = str_replace_all(gseTest_h$pathway, "_", " ") %>%
  str_replace(., paste0(geneSet, " "), "") #%>%
# str_to_title()

#first make enrichment score plots (these need to be included in these code chunks because the size of the text changes)

pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '.pdf'), height=10, width=10)

ggplot(gseTest_h, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip()+
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 4))

#make some plots, but only include pathways with padj<0.05
ggplot(gseTest_h[gseTest_h$padj<0.05,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 10))+
  scale_fill_manual(values="#355C7D")

#only plots w/ padj <0.05 and NES > 0
ggplot(gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,],aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 10))+
  scale_fill_manual(values="#355C7D")
dev.off()
#then run stock code
<<gsea_plots>>
```

#### GSEA for WIKIPATHWAYS gene sets (n=712)

```{r WIKIPATHWAYS, echo=FALSE}
geneSet = "WP" #set name of gene set under analysis
m_df = msigdbr(species = "Homo sapiens", category = 'C2', subcategory = "CP:WIKIPATHWAYS") # gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name) 
names(m_list) = str_replace_all(names(m_list),"_"," ") %>%    str_replace(., paste0(geneSet, " "), "") #%>%

#gsea_run block is commented out since I've run it before, instead load the pathwayAnalysisResults file
<<gsea_run>>
# gseTest_h <- read_csv(file = paste0(output.path, geneSet, "/pathwayAnalysisResults_", geneSet, ".csv"), show_col_types=F)

##############
#make gsea plots
#clean up names on plots
gseTest_h$pathway = str_replace_all(gseTest_h$pathway, "_", " ") %>%
  str_replace(., paste0(geneSet, " "), "") #%>%
# str_to_title()

#first make enrichment score plots (these need to be included in these code chunks because the size of the text changes)

pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '.pdf'), height=10, width=10)

ggplot(gseTest_h, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip()+
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 2))

#make some plots, but only include pathways with padj<0.05
ggplot(gseTest_h[gseTest_h$padj<0.05,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 4))+
  scale_fill_manual(values="#355C7D")

#only plots w/ padj <0.05 and NES > 0
ggplot(gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,],aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 8))+
  scale_fill_manual(values="#355C7D")

#make same plots, but only include pathways with padj<0.05 and subset for top and bottom 10 by NES
for(number in c(10, 20, 30, 40, 50)) {
  df <- gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,] %>% top_n(.,number,wt=NES)
  p <- ggplot(df,aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0("Top ", number, " ", geneSet," pathways")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 10))+
  scale_fill_manual(values="#355C7D") 
print(p)
rm(df)
rm(p)
}
dev.off()

#then run stock code
<<gsea_plots>>
```

#### GSEA for REACTOME gene sets (n=1635)

```{r REACTOME, echo=FALSE}
geneSet = "REACTOME" #set name of gene set under analysis
m_df = msigdbr(species = "Homo sapiens", category = 'C2', subcategory = "CP:REACTOME") # gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name) 
names(m_list) = str_replace_all(names(m_list),"_"," ") %>%    str_replace(., paste0(geneSet, " "), "") #%>%

#gsea_run block is commented out since I've run it before, instead load the pathwayAnalysisResults file
<<gsea_run>>
# gseTest_h <- read_csv(file = paste0(output.path, geneSet, "/pathwayAnalysisResults_", geneSet, ".csv"), show_col_types=F)

##############
#make gsea plots
#clean up names on plots
gseTest_h$pathway = str_replace_all(gseTest_h$pathway, "_", " ") %>%
  str_replace(., paste0(geneSet, " "), "") #%>%
# str_to_title()

#first make enrichment score plots (these need to be included in these code chunks because the size of the text changes)

pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '.pdf'), height=10, width=10)

ggplot(gseTest_h, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip()+
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 0.5))

#make some plots, but only include pathways with padj<0.05
ggplot(gseTest_h[gseTest_h$padj<0.05,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 1))+
  scale_fill_manual(values="#355C7D")

#only plots w/ padj <0.05 and NES > 0
ggplot(gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,],aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 8))+
  scale_fill_manual(values="#355C7D")

#make same plots, but only include pathways with padj<0.05 and subset for top and bottom 10 by NES
for(number in c(10, 20, 30, 40, 50)) {
  df <- gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,] %>% top_n(.,number,wt=NES)
  p <- ggplot(df,aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0("Top ", number, " ", geneSet," pathways")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 10))+
  scale_fill_manual(values="#355C7D") 
print(p)
rm(df)
rm(p)
}
dev.off()

#then run stock code
<<gsea_plots>>
```

#### GSEA for GOBP gene sets (n=7763)

```{r GOBP, echo=FALSE}
geneSet = "GOBP" #set name of gene set under analysis
m_df = msigdbr(species = "Homo sapiens", category = 'C5', subcategory = "GO:BP") # gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name)  
names(m_list) = str_replace_all(names(m_list),"_"," ") %>%    str_replace(., paste0(geneSet, " "), "") #%>%

#gsea_run block is commented out since I've run it before, instead load the pathwayAnalysisResults file
# <<gsea_run>>
gseTest_h <- read_csv(file = paste0(output.path, geneSet, "/pathwayAnalysisResults_", geneSet, ".csv"), show_col_types=F)

##############
#make gsea plots
#clean up names on plots
gseTest_h$pathway = str_replace_all(gseTest_h$pathway, "_", " ") %>%
  str_replace(., paste0(geneSet, " "), "") #%>%
# str_to_title()

#first make enrichment score plots (these need to be included in these code chunks because the size of the text changes)

pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '.pdf'), height=10, width=10)

ggplot(gseTest_h, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip()+
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 0.01))

#make some plots, but only include pathways with padj<0.05
ggplot(gseTest_h[gseTest_h$padj<0.05,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 0.5), legend.position="null")+
  scale_fill_manual(values="#355C7D")

#only plots w/ padj <0.05 and NES > 0
ggplot(gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,],aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 1))+
  scale_fill_manual(values="#355C7D")

#make same plots, but only include pathways with padj<0.05 and subset for top and bottom 10 by NES
for(number in c(10, 20, 30, 40,50)) {
  df <- gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,] %>% top_n(.,number,wt=NES)
  p <- ggplot(df,aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0("Top ", number, " ", geneSet," pathways")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 8))+
  scale_fill_manual(values="#355C7D") 
print(p)
rm(df)
rm(p)
}
dev.off()
#then run stock code
<<gsea_plots>>


### generate plots for enrichment score

########
########
#make object up.pathways for every gene set identified with p < 0.05 and positive enrichment score
gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES>0,]$pathway -> up.pathways
pathways.of.interest <- c("T CELL ACTIVATION", "B CELL ACTIVATION","DENDRITIC CELL ANTIGEN PROCESSING AND PRESENTATION")
pathways.of.interest %in% up.pathways

library(fgsea)
stat <- DE_res$stat
names(stat) <- row.names(DE_res)

plot_list <- lapply(1:length(pathways.of.interest), function(i) { 
  pathway = pathways.of.interest[i]
 x <- plotEnrichment(m_list[[pathway]],stat)+ 
   # labs(title = str_wrap(pathway, 40))+
   theme(plot.title =element_text(hjust=0.5,face="bold",size=16),
         axis.title = element_text(face="bold",size=14),
         axis.text = element_text(face="bold",size=12))
# x$layers[[1]]$aes_params$size  = 1 #this comes from https://stackoverflow.com/questions/64035941/changing-the-point-size-of-a-ggplot2-object-after-creation
  x$layers[[5]]$aes_params$size = 1
  return(x)
}) %>% `names<-`(pathways.of.interest)
plot_list$`DENDRITIC CELL ANTIGEN PROCESSING AND PRESENTATION`$labels$title <- "DENDRITIC CELL ANTIGEN PROCESSING \nAND PRESENTATION"

#plot these
pdf(paste0(output.path, geneSet, "/gsea_enrichment_score_",geneSet,"_3_pathwaysOfInterest.pdf"),
    width=7,height=5
    )
plot_list$`T CELL ACTIVATION`
plot_list$`B CELL ACTIVATION`
plot_list$`DENDRITIC CELL ANTIGEN PROCESSING AND PRESENTATION`
dev.off()
```

#### GSEA for GOCC gene sets (n=1035)

```{r GOCC, echo=FALSE}
geneSet = "GOCC" #set name of gene set under analysis
m_df = msigdbr(species = "Homo sapiens", category = 'C5', subcategory = "GO:CC") # gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name) 
names(m_list) = str_replace_all(names(m_list),"_"," ") %>%    str_replace(., paste0(geneSet, " "), "") #%>%

#gsea_run block is commented out since I've run it before, instead load the pathwayAnalysisResults file
<<gsea_run>>
# gseTest_h <- read_csv(file = paste0(output.path, geneSet, "/pathwayAnalysisResults_", geneSet, ".csv"), show_col_types=F)

##############
#make gsea plots
#clean up names on plots
gseTest_h$pathway = str_replace_all(gseTest_h$pathway, "_", " ") %>%
  str_replace(., paste0(geneSet, " "), "") #%>%
# str_to_title()

#first make enrichment score plots (these need to be included in these code chunks because the size of the text changes)

pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '.pdf'), height=10, width=10)

ggplot(gseTest_h, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip()+
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 0.01))

#make some plots, but only include pathways with padj<0.05
ggplot(gseTest_h[gseTest_h$padj<0.05,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 1))+
  scale_fill_manual(values="#355C7D")

#only plots w/ padj <0.05 and NES > 0
ggplot(gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,],aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 6))+
  scale_fill_manual(values="#355C7D")

#make same plots, but only include pathways with padj<0.05 and subset for top and bottom 10 by NES
for(number in c(10, 20, 30, 40,50)) {
  df <- gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,] %>% top_n(.,number,wt=NES)
  p <- ggplot(df,aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0("Top ", number, " ", geneSet," pathways")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 8))+
  scale_fill_manual(values="#355C7D") 
print(p)
rm(df)
rm(p)
}
dev.off()

#then run stock code
<<gsea_plots>>

```

#### GSEA for GOMF gene sets (n=1763)

```{r GOMF, echo=FALSE}
geneSet = "GOMF" #set name of gene set under analysis
m_df = msigdbr(species = "Homo sapiens", category = 'C5', subcategory = "GO:MF") # gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name)  
names(m_list) = str_replace_all(names(m_list),"_"," ") %>%    str_replace(., paste0(geneSet, " "), "") #%>%

#gsea_run block is commented out since I've run it before, instead load the pathwayAnalysisResults file
<<gsea_run>>
# gseTest_h <- read_csv(file = paste0(output.path, geneSet, "/pathwayAnalysisResults_", geneSet, ".csv"), show_col_types=F)

##############
#make gsea plots
#clean up names on plots
gseTest_h$pathway = str_replace_all(gseTest_h$pathway, "_", " ") %>%
  str_replace(., paste0(geneSet, " "), "") #%>%
# str_to_title()

#first make enrichment score plots (these need to be included in these code chunks because the size of the text changes)

pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '.pdf'), height=10, width=10)

ggplot(gseTest_h, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip()+
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 0.01))

#make some plots, but only include pathways with padj<0.05
ggplot(gseTest_h[gseTest_h$padj<0.05,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 2))+
  scale_fill_manual(values="#355C7D")

#only plots w/ padj <0.05 and NES > 0
ggplot(gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,],aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 1))+
  scale_fill_manual(values="#355C7D")

#make same plots, but only include pathways with padj<0.05 and subset for top and bottom 10 by NES
for(number in c(10, 20, 30, 40,50)) {
  df <- gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,] %>% top_n(.,number,wt=NES)
  p <- ggplot(df,aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0("Top ", number, " ", geneSet," pathways")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 8))+
  scale_fill_manual(values="#355C7D") 
print(p)
rm(df)
rm(p)
}
dev.off()
#then run stock code
<<gsea_plots>>
```

#### GSEA for IMMUNESIGDB gene sets

```{r IMMUNESIGDB, eval=F}
geneSet = "IMMUNESIGDB" #set name of gene set under analysis
m_df = msigdbr(species = "Homo sapiens", category = 'C7', subcategory = "IMMUNESIGDB") # gene sets for pathway analysis
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name) 
names(m_list) = str_replace_all(names(m_list),"_"," ")# %>%    str_replace(., paste0(geneSet, " "), "") #%>%

#gsea_run block is commented out since I've run it before, instead load the pathwayAnalysisResults file
<<gsea_run>>
# gseTest_h <- read_csv(file = paste0(output.path, geneSet, "/pathwayAnalysisResults_", geneSet, ".csv"), show_col_types=F)

##############
#make gsea plots
#clean up names on plots
gseTest_h$pathway = str_replace_all(gseTest_h$pathway, "_", " ")# %>%
  # str_replace(., paste0(geneSet, " "), "") #%>%
# str_to_title()

#first make enrichment score plots (these need to be included in these code chunks because the size of the text changes)

pdf(file = paste0(output.path, geneSet, '/gsea_NES_', geneSet, '.pdf'), height=10, width=10)

ggplot(gseTest_h, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip()+
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 0.01))

#make some plots, but only include pathways with padj<0.05
ggplot(gseTest_h[gseTest_h$padj<0.05,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
       ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 0.5))+
  scale_fill_manual(values="#355C7D")

#only plots w/ padj <0.05 and NES > 0
ggplot(gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,],aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title= paste0(geneSet," pathways NES from GSEA")
  ) + 
  theme_minimal()+
  theme(axis.text.y = element_text(size = 1))+
  scale_fill_manual(values="#355C7D")

#make same plots, but only include pathways with padj<0.05 and subset for top and bottom 10 by NES
for(number in c(10, 20, 30, 40,50)) {
  df <- gseTest_h[gseTest_h$padj<0.05 & gseTest_h$NES >0,] %>% top_n(.,number,wt=NES)
  p <- ggplot(df,aes(reorder(pathway, NES), NES)) +
    geom_col(aes(fill=padj<0.05)) +
    coord_flip() +
    labs(x="Pathway", y="Normalized Enrichment Score",
         title= paste0("Top ", number, " ", geneSet," pathways")
    ) + 
    theme_minimal()+
    theme(axis.text.y = element_text(size = 8))+
    scale_fill_manual(values="#355C7D") 
  print(p)
  rm(df)
  rm(p)
}
dev.off()
#then run stock code
<<gsea_plots>>
```

# III. Collect as xlsx
```{r}
files <- list.files(path=output.path, pattern="pathwayAnalysisResults_", recursive=T, full.names = T)
list <- lapply(files, FUN=read.csv)
names(list)  = files %>% str_replace("^.*[_]","") %>% str_replace(".csv", "")

writexl::write_xlsx(list, path=paste0(output.path,"pathwayAnalysisResultsCombined.xlsx"))

```

# IV. SessionInfo

```{r echo=FALSE}
sessionInfo()
writeLines(capture.output(sessionInfo()), "sessionInfo_BulkRNAseq_GSEA.txt")
```
