---
title: "Metadata creation"
author: "Daniel Shu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# I. Load libraries

```{r}
library(tidyverse)
library(readxl)
library(immunarch)
library(kableExtra)

data_dir = "data/adaptive/"
output.path = "output/adaptive/"
#creates output.path directory if not already present
ifelse(!dir.exists(output.path), dir.create(output.path), paste0(output.path, " ", "directory already exists"))
```

# II. Immunarch
### a. Create and save HCC_immdata object
note: these were downloaded from the adaptive website after filtering for in frame rearrangements, i.e. there are no noncoding clonotypes
```{r}
HCC_immdata <- repLoad(.path=data_dir)

# HCC_immdata <- readRDS(file=paste0(output.path,"3b_compiled_HCC_immdata_obj.rds"))
# saveRDS(HCC_immdata, file=paste0(output.path,"3b_compiled_HCC_immdata_obj.rds"))

```

### b. Set Levels
```{r}
HCC_immdata$meta$time_point <- factor(HCC_immdata$meta$time_point, levels=c("Pre", "Post"))
HCC_immdata$meta$repertoire <- factor(HCC_immdata$meta$repertoire, levels=c("TCRB", "IGH"))
HCC_immdata$meta$TLS_type <- factor(HCC_immdata$meta$TLS_type, levels=c("Mature", "Involuted"))
HCC_immdata$meta$Location <- factor(HCC_immdata$meta$Location, levels=c("Intratumoral", "Peritumoral", "Adjacent Normal"))
HCC_immdata$meta$Confirmed_CD21 <- factor(HCC_immdata$meta$Confirmed_CD21, levels=c("Yes", "No"), labels=c("CD21pos", "CD21neg"))
```


### c. Modify metadata by adding additional metadata pertaining to total # of clones (Clones), unique # of clones (Volume), and Shannon entropy
Shannon Entropy code based on code from Janelle Montagne
#### 1. total clones and unique clones 
```{r}
#adds # of total clones per sample to metadata and plots
p1 <- repExplore(HCC_immdata$data, .method = "clones") #number of clones total
p1
HCC_immdata$meta <- merge(HCC_immdata$meta, p1, by = "Sample")

#adds # of # of unique clones per sample to metadata and plots
p2 <- repExplore(HCC_immdata$data, .method = "volume") #number of unique clonotypes
HCC_immdata$meta <- merge(HCC_immdata$meta, p2, by = "Sample")
p2
```

#### 2. shannon entropy

```{r}
#### Add shannon entropy
#add Clonality column for Normalized Shannon's Entropy (Shannon's equitability- measure of evenness)
#between 0 and 1, lower value indicates more evenness (more polyclonal), higher value indicates less evenness (1 is monoclonal)

div_div <- repDiversity(coding(
  HCC_immdata$data)
  , "div", .q = 1, .col ="aa"
) #.q = 1 gives you exp shannon
div_div$clonotypes <- 0
#convert from true diversity back to Shannon entropy
div_div$Value <- log(div_div$Value)
#Get number of clonotypes per sample to calculate normalized Shannon entropy
for (i in 1:length(HCC_immdata[[1]])){
  sample <- names(HCC_immdata[[1]][i])
  test <- data.frame(coding(HCC_immdata[[1]][i]))
  clonotypes <- length(unique(test[,4]))##make it based on amino acid clonotypes
  div_div$clonotypes[i] <- clonotypes
}

# div_div$clonotypes <- div_div$clonotypes + 0.1 #to avoid 0 values
#Calculate normalized Shannon entropy
div_div$lnClonotypes <- log(div_div$clonotypes)
div_div$shannon_norm <- div_div$Value/div_div$lnClonotypes
div_div$Value <- 1-div_div$shannon_norm

##Assign clonality to metadata
clonalityDF <- div_div[,c(1:2)]
colnames(clonalityDF) <- c("Sample", "clonality")
HCC_immdata$meta <- merge(HCC_immdata$meta, clonalityDF, by = "Sample")
```

#### 3. Add column if TCR was present in pre or post treatment PBMC

```{r}
patients = HCC_immdata$meta$Patient.ID %>% unique

#for all patients, adds columsn if the tcr in specific tls sampels was present in the pre-treatment biopsy
for (x in 1:length(patients)){
  patient = patients[x]
  samples.of.interest = HCC_immdata$meta[HCC_immdata$meta$Patient.ID == patient & HCC_immdata$meta$repertoire=="TCRB" & HCC_immdata$meta$type=="TLS",]$Sample
  # biopsy.of.interest = HCC_immdata$meta[HCC_immdata$meta$Patient.ID == patient & HCC_immdata$meta$type=="Biopsy",]$Sample
  pbmc.pre.of.interest = HCC_immdata$meta[HCC_immdata$meta$Patient.ID ==patient & HCC_immdata$meta$time_point=="Pre" & HCC_immdata$meta$type=="PBMC",]$Sample
  pbmc.post.of.interest = HCC_immdata$meta[HCC_immdata$meta$Patient.ID ==patient & HCC_immdata$meta$time_point=="Post" & HCC_immdata$meta$type=="PBMC",]$Sample

  for (i in 1:length(samples.of.interest)){ 
    # if (is_empty(biopsy.of.interest) == F) {HCC_immdata$data[[samples.of.interest[i]]]$present.in.biopsy =
    #   if_else(HCC_immdata$data[[samples.of.interest[i]]]$CDR3.nt%in%HCC_immdata$data[[biopsy.of.interest]]$CDR3.nt, 1, 0)
    # } else {
    #   print(paste0("no biopsy sample for patient ",patient, " so present.in.biopsy column not added for sample ", samples.of.interest[i]))
    #   HCC_immdata$data[[samples.of.interest[i]]]$present.in.biopsy = NA
    #   }
    #then if the patients have pre-post pbmc, add column to samples.of.interest, i.e. TLS dataframes, denoting if present in pre or post pbmc
  if (patient %in% patients[1:5]){
    HCC_immdata$data[[samples.of.interest[i]]]$present.in.pbmc.pre = if_else(HCC_immdata$data[[samples.of.interest[i]]]$CDR3.nt%in%HCC_immdata$data[[pbmc.pre.of.interest]]$CDR3.nt, 1, 0)
    HCC_immdata$data[[samples.of.interest[i]]]$present.in.pbmc.post = if_else(HCC_immdata$data[[samples.of.interest[i]]]$CDR3.nt%in%HCC_immdata$data[[pbmc.post.of.interest]]$CDR3.nt, 1, 0)}
  else {
    HCC_immdata$data[[samples.of.interest[i]]]$present.in.pbmc.pre = NA
    HCC_immdata$data[[samples.of.interest[i]]]$present.in.pbmc.post = NA}
  }
}
```

#### 4. For B cell repertoires, add valignsubstitution column

```{r eval = T}
bcr_adaptive_files <- list.files(path="data/adaptive", pattern = ".*-B.tsv",full.names=T)
bcr_adaptive_files
bcr_adaptive_data <- lapply(1:length(bcr_adaptive_files), function(i) {
  read_tsv(file=bcr_adaptive_files[i],show_col_types = F)})
names(bcr_adaptive_data) = bcr_adaptive_files %>% str_replace(.,".*adaptive/","") %>% str_replace(".tsv","")
names(bcr_adaptive_data)
bcr_adaptive_data_trimmed <- bcr_adaptive_data %>% bind_rows(.id="sample")  %>% 
  select(sample, nucleotide,vAlignSubstitutionCount) %>% 
  rename_with(.,tolower) %>%    #change all names to lowercase
  split(., .$sample)   
if (all(names(bcr_adaptive_data_trimmed) %in% names(HCC_immdata$data))==T){ #doublecheck that names are present
  bnames = names(bcr_adaptive_data_trimmed)
  for (i in bnames) {
    HCC_immdata$data[[i]] = left_join(HCC_immdata$data[[i]],
                                      select(bcr_adaptive_data_trimmed[[i]],nucleotide,valignsubstitutioncount),
                                      by=join_by(Sequence==nucleotide))
  }
} 

```

### d. Save object 
```{r}
# Save object for further analysis
saveRDS(HCC_immdata, paste0(output.path, "HCC_immdata.rds"))
# HCC_immdata <- readRDS(paste0(output.path, "HCC_immdata.rds")) #this line loads the object
```

### e. Export table

```{r}
write.csv(HCC_immdata$meta, paste0(output.path, "HCC_immdata_metadata_all.csv"), row.names=F)
```
# III. SessionInfo
```{r}
sessionInfo()
writeLines(capture.output(sessionInfo()), "sessionInfo_ImmunoSEQ_load_immunarchObject_and_filter.txt")
```


